\documentclass{article}

\usepackage[english]{babel}
\usepackage[letterpaper,top=2cm,bottom=2cm,left=2cm,right=2cm]{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds}
\setlength{\parindent}{0pt}

\title{Milestone 1}
\author{Eunice Chen, Muhaimen Shamsi, Dariia Kucheruk, Oleksii Nakhod}
\date{CSC2541H / PCL3107H / PCL3108H, Winter 2026}

\begin{document}
\maketitle

\section*{Part I: Introduction and Background (1 Page Maximum)}

\subsection*{1. Research Question \& Significance}
\textbf{Research question} Do AD-implicated genes show stronger and more specific signal than non-AD controls in (i) DrugCLIP embedding neighborhoods and (ii) independent PPI topology, and can this be used to prioritize plausible drug candidates?

\textbf{Why this matters for AI in drug discovery} Drug discovery embedding models are usually evaluated on benchmark task performance, but not always on disease-specific biological validity. Our goal is to test whether embedding-derived structure is aligned with AD biology rather than generic signal. If yes, this supports using embedding + network evidence for disease-focused target prioritization.

\subsection*{2. Context \& Datasets}
\textbf{Context} We combine representation-learning diagnostics and network medicine validation. We use matched controls and degree-matched nulls to avoid inflated claims from graph hubs.

\textbf{Datasets used}
\begin{itemize}
    \item DrugCLIP protein embeddings and protein names from THU-ATOM DrugCLIP data \cite{adGwas2025, drugclipData2024}.
    \item AD gene list derived from the Askenazi 2023 supplement.
    \item Allen Human Brain RNA-seq donor 9861 bulk expression release, used for expression-based AD/control label construction \cite{allenAtlas2012}.
    \item HGNC complete symbol mapping table for gene-symbol to UniProt harmonization \cite{hgncCompleteSet}.
    \item STRING v12 human protein metadata and interaction tables, used to build the PPI graph for topology and degree-matched controls \cite{string2025}.
    \item DrugCLIP target assay archive for candidate-drug retrieval.
\end{itemize}

\textbf{Current status before Milestone 2} The latest multiseed workflow run supports a real AD-vs-null signal in held-out probe metrics and in independent PPI topology. Observed AD beats both label-shuffle and matched random-set baselines on AUROC/AUPRC and shows stronger PPI concentration and cosine-neighbor AD enrichment. The main remaining gap is that the random-embedding ablation is not yet fully propagated through the reported network summaries, so that control still needs a clean end-to-end validation pass.

\section*{Part II: Proposed Methods and Setup (2 Pages)}

\subsection*{3. Proposed Methods}
\textbf{Method A: AD-vs-control prediction head on embeddings} We aggregate multiple embedding rows to one gene vector and train a small FFN (768$\rightarrow$64$\rightarrow$1). We compare three ablations: embedding, random-embedding, and label-shuffle.

\textbf{Method B: Embedding-to-network biological validation} We test whether AD labels are topologically concentrated in PPI with degree-matched permutation nulls. Metrics include AD-AD edge enrichment, AD largest connected component, and AD-neighbor fraction.

\textbf{Method C: Cosine-neighbor/PPI coupling by hop depth} For $k\in\{1,2,5\}$ cosine neighbors, we compare AD-seed vs control-seed neighbor properties (AD fraction and PPI coupling), then evaluate significance with degree-matched permutation.

\textbf{Method D: Candidate drug retrieval and characterization} We retrieve potent compounds from DrugCLIP ChEMBL target tables for AD-overlap targets (with alias mapping), rank compounds by AD-target coverage and potency support, then characterize phase/potency/target distributions.

\textbf{Innovation claim} The novelty is not a new foundation model; it is a disease-specific validation stack that links embedding geometry to independent biological topology using strict controls (label-shuffle, random-embedding, degree-matched nulls).

\subsection*{4. Data and Preprocessing}
\textbf{ID harmonization} Gene symbols are standardized to uppercase and mapped to UniProt accessions (HGNC mapping) for embedding joins.

\textbf{Leakage prevention}
\begin{itemize}
    \item Aggregation to one embedding per gene is done \emph{before} train/test splitting.
    \item Splits are fixed by saved split files for fair ablation comparison.
    \item PPI analyses use degree-matched permutation controls to reduce hub-driven artifacts.
\end{itemize}

\textbf{Filtering and thresholds}
\begin{itemize}
    \item PPI analyses use STRING combined score threshold (default 700 for high confidence).
    \item Drug retrieval keeps potency-supported rows (default: pChEMBL $\geq 6$ or standard value $\leq$ 1000 nM where applicable).
\end{itemize}

\subsection*{5. Information Flow Diagram}
\begin{figure}[h]
    \centering
    \resizebox{\linewidth}{!}{%
        \begin{tikzpicture}[
            font=\sffamily,
            >=Latex,
            node distance=7mm and 8mm,
            io/.style={draw, rounded corners=3pt, line width=0.8pt, minimum width=3.4cm, minimum height=0.95cm, align=center, fill=#1, fill opacity=1, text opacity=1},
            stage/.style={draw, rounded corners=4pt, line width=1pt, minimum width=3.9cm, minimum height=1.2cm, align=center, fill=#1, fill opacity=1, text opacity=1},
            out/.style={draw, rounded corners=3pt, line width=0.8pt, minimum width=3.5cm, minimum height=0.95cm, align=center, fill=#1, fill opacity=1, text opacity=1},
            arrow/.style={-{Latex[length=2.5mm]}, line width=0.9pt, draw=black!75, rounded corners=4pt},
            faint/.style={draw=black!35, dashed, rounded corners=6pt}
            ]

            % Inputs
            \node[io=blue!8] (ad) {\textbf{AD seed genes}\\Askenazi supplement};
            \node[io=blue!8, below=of ad] (rna) {\textbf{Brain expression}\\Allen RNA-seq};
            \node[io=blue!8, below=of rna] (hgnc) {\textbf{ID crosswalk}\\HGNC $\rightarrow$ UniProt};

            \node[io=green!9, right=28mm of rna] (emb) {\textbf{Protein embeddings}\\DrugCLIP vectors + names};
            \node[io=green!9, above=of emb] (ppi) {\textbf{Biological network}\\STRING v12 PPI};
            \node[io=green!9, below=of emb] (chembl) {\textbf{Compound evidence}\\DrugCLIP assay archive};

            % Central preprocessing
            \node[stage=orange!11, right=26mm of emb] (labels) {\textbf{Label construction}\\AD vs matched control genes\\regional expression filtering};
            \node[stage=orange!11, below=of labels] (join) {\textbf{Gene--protein join}\\aggregate embedding rows\\to one vector per gene};

            % Analysis row
            \node[stage=purple!10, right=24mm of labels] (probe) {\textbf{Prediction probe}\\small classifier head\\AUROC, AUPRC, accuracy};
            \node[stage=purple!10, right=24mm of join] (hops) {\textbf{Embedding / PPI coupling}\\cosine neighbors at $k=1,2,5$\\degree-matched nulls};
            \node[stage=teal!10, below=of hops] (topology) {\textbf{Network validation}\\AD-AD enrichment, LCC,\\AD-neighbor fraction};

            % Output
            \node[out=yellow!14, right=23mm of hops] (summary) {\textbf{Evidence bundle}\\embedding signal\\+ topology support\\+ ranked outputs};
            \node[out=yellow!14, above=of summary] (controls) {\textbf{Controls / ablations}\\label shuffle\\random embedding\\random matched set};
            \node[out=yellow!14, below=of summary] (drugs) {\textbf{Candidate retrieval}\\potent compounds for\\AD-overlap targets};

            % Routing coordinates
            \coordinate[right=9mm of hgnc.east] (c-input-low);
            \coordinate[right=9mm of rna.east] (c-input-mid);
            \coordinate[right=9mm of ad.east] (c-input-high);
            \coordinate[right=9mm of emb.east] (c-emb-out);
            \coordinate[right=10mm of join.east] (c-join-out);
            \coordinate[right=10mm of ppi.east] (c-ppi-out);
            \coordinate[left=14mm of summary.west] (c-summary-in);
            \coordinate[above left=7mm and 14mm of summary.west] (c-summary-in-high);
            \coordinate[below left=7mm and 14mm of summary.west] (c-summary-in-low);

            % Group boxes
            \begin{scope}[on background layer]
                % Arrows
                \draw[arrow] (ad.east) -- (c-input-high) |- (labels.west);
                \draw[arrow] (rna.east) -- (c-input-mid) |- (labels.west);
                \draw[arrow] (hgnc.east) -- (c-input-low) |- (join.west);
                \draw[arrow] (emb.east) -- ++(8mm,0) |- (join.west);
                \draw[arrow] (emb.east) -- (c-emb-out) |- ([yshift=4pt]hops.west);
                \draw[arrow] (labels.south) -- (join.north);
                \draw[arrow] (ppi.east) -- (c-ppi-out) |- ([yshift=8pt]hops.north);
                \draw[arrow] (ppi.east) -- (c-ppi-out) |- ([yshift=6pt]topology.north);
                \draw[arrow] (chembl.east) -- ++(10mm,0) |- (drugs.west);

                \draw[arrow] (join.east) -- (c-join-out) |- ([yshift=-3pt]probe.west);
                \draw[arrow] (join.east) -- (c-join-out) |- (hops.west);
                \draw[arrow] (join.east) -- (c-join-out) |- ([yshift=2pt]topology.west);

                \draw[arrow] (probe.east) -- (c-summary-in-high) -- ([yshift=8pt]summary.west);
                \draw[arrow] (hops.east) -- (c-summary-in) -- (summary.west);
                \draw[arrow] (topology.east) -- (c-summary-in-low) -- ([yshift=-8pt]summary.west);
                \draw[arrow] (summary.south) -- (drugs.north);
                \draw[arrow] (controls.south) -- (summary.north);

                \node[faint, fit=(ad)(rna)(hgnc)(emb)(ppi)(chembl), inner sep=8pt, label={[font=\bfseries\sffamily]above:Input resources}] {};
                \node[faint, fit=(labels)(join), inner sep=8pt, label={[font=\bfseries\sffamily]above:Preprocessing and alignment}] {};
                \node[faint, fit=(probe)(hops)(topology), inner sep=8pt, label={[font=\bfseries\sffamily]above:Analysis modules}] {};
                \node[faint, fit=(controls)(summary)(drugs), inner sep=8pt, label={[font=\bfseries\sffamily]above:Reported outputs}] {};
            \end{scope}
        \end{tikzpicture}%
    }
    \caption{Architecture and information flow used in Milestone 1 experiments}
\end{figure}

\subsection*{6. Experimental Setup \& Baselines}
\textbf{Primary metrics}
\begin{itemize}
    \item Predictor: test AUROC, AUPRC, accuracy (multi-seed mean/stdev).
    \item PPI topology: AD-AD enrichment, AD LCC size, AD-neighbor fraction with permutation p-values.
    \item Hop analysis: AD-control delta at $k=1,2,5$ with degree-matched permutation p-values.
\end{itemize}

\textbf{Baselines locked in}
\begin{itemize}
    \item Label-shuffle baseline.
    \item Matched random non-AD set baseline.
    \item Degree-matched permutation null for PPI and hop analyses.
    \item Random-embedding ablation in the probe pipeline; full network-level validation still being finalized.
    \item Seed policy: 10 seeds for routine development/debugging runs, 30 seeds for final reported stability estimates.
\end{itemize}

\textbf{Current quantitative snapshot}
\begin{itemize}
    \item In the current multiseed bundle, observed AD probe performance is above both null baselines: mean AUROC $=0.533$ and mean AUPRC $=0.461$ versus $\approx 0.501$ AUROC and $\approx 0.345$ AUPRC for both label-shuffle and matched random-set baselines.
    \item PPI topology remains the strongest signal: observed AD-AD edge enrichment is $\approx 1.29\times$ versus $\approx 0.99\times$ under nulls; AD largest connected component size is $605$ versus $\approx 419$ under nulls; mean AD-neighbor fraction is $0.439$ versus $\approx 0.332$ under nulls.
    \item Cosine/PPI hop analyses are directionally consistent with AD-specific local structure: AD-neighbor fraction deltas are positive at $k=1,2,5$ ($0.103$, $0.098$, $0.052$) and are near zero under the null baselines.
    \item Hypothesis testing on the strongest topology and hop metrics gives empirical $p \approx 0.024$ but BH-FDR $q \approx 0.062$, so the current evidence is strong and consistent but still just above a strict $q<0.05$ cutoff.
    \item After improving synonym and composite target-name resolution, the number of AD-relevant targets with retrievable compound evidence increased from 2 to 5.
    \item The current multiseed experiment bundle took approximately 2 hours and 12 minutes end-to-end, using development-scale settings of 10 seeds and 100 degree-matched permutations rather than a larger final-reporting run.
\end{itemize}

\subsection*{7. Timeline to Milestone 2}
\textbf{Week of March 8--14, 2026}
\begin{itemize}
    \item TODO
    \item TODO
\end{itemize}

\textbf{Week of March 15--21, 2026}
\begin{itemize}
    \item TODO
    \item TODO
\end{itemize}

\textbf{Week of March 22--28, 2026}
\begin{itemize}
    \item TODO
    \item TODO
\end{itemize}

\textbf{Week of March 29--April 4, 2026}
\begin{itemize}
    \item TODO
    \item TODO
\end{itemize}

\textbf{Anticipated bottlenecks}
\begin{itemize}
    \item Long permutation jobs on large PPI graphs.
    \item Runtime variability from large table parsing and multi-seed reruns.
    \item Limited AD-target overlap in current DrugCLIP target archive.
\end{itemize}

\bibliographystyle{unsrturl}
\bibliography{milestone1}

\end{document}
