\documentclass{article}

% Language setting
\usepackage[english]{babel}

% Set page size and margins
\usepackage[letterpaper,top=2cm,bottom=2cm,left=2cm,right=2cm,marginparwidth=1.75cm]{geometry}

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\setlength{\parindent}{0pt}

\title{Letter of Intent}
\author{Eunice Chen, Muhaimen Shamsi, Dariia Kucheruk, Oleksii Nakhod}

\begin{document}
\maketitle

\section{Research Question}
Do AD-implicated genes (from AD GWAS/TWAS) show the strongest and most specific
signals in AD-vulnerable brain regions and related network neighborhoods, compared
with matched non-AD control genes?

\section{Significance}
DrugCLIP works well for virtual screening at scale \cite{adGwas2025}. The open question is whether its embeddings capture AD-relevant biology, not just generic screening signal. We therefore focus disease inputs strictly on AD-implicated genes and test whether they are enriched in AD-vulnerable regions (entorhinal cortex, hippocampus, and temporal cortex) and nearby PPI neighborhoods \cite{aiNetworkMedicine2025}. We also include a matched non-AD control gene set and run the exact same scoring pipeline to test specificity. A positive result means the signal is AD-specific and biologically grounded, which can improve target prioritization.

\section{Current State of the Field}
Most embedding papers report task metrics (for example enrichment in screening benchmarks). Network medicine papers, on the other hand, evaluate disease modules and graph proximity in PPI networks \cite{string2018, string2019}. What is often missing is a disease-specific control framework. Here we bridge these areas by evaluating AD genes against matched non-AD controls under the same pipeline, with explicit emphasis on AD-vulnerable regions.

\section{Existing Models}
Current approaches can be grouped as:
\begin{itemize}
    \item \textbf{Task-focused embedding models} optimized for virtual screening performance \cite{adGwas2025}.
    \item \textbf{Network medicine methods} based on disease gene sets and PPI proximity \cite{string2018}.
    \item \textbf{Expression-centric approaches} using region or cell-type profiles \cite{allenAtlas2012, seaad2024}.
\end{itemize}
Our project does one thing clearly: test whether AD-implicated genes receive stronger AD-region and network support than non-AD controls under a unified scoring framework.

\section{Available Datasets}
\begin{itemize}
    \item \textbf{DrugCLIP protein embeddings}: pretrained vectors for proteins from published DrugCLIP resources \cite{adGwas2025, drugclipData2024}.
    \item \textbf{AD disease gene set (primary input)}: AD-implicated genes from AD GWAS/TWAS resources.
    \item \textbf{Control gene set (negative control)}: non-AD genes matched on key properties (for example expression level and network degree), optionally including atlas-derived stable or housekeeping-like genes.
    \item \textbf{PPI network}: high-confidence protein interaction graph used for module and proximity tests \cite{string2018, string2019}.
    \item \textbf{Region/cell-type expression metadata}: Allen/SEA-AD resources, with primary focus on entorhinal cortex, hippocampus, and temporal cortex \cite{allenAtlas2012, seaad2024}.
\end{itemize}

\section{Data Generation}
No new wet-lab data will be generated in this LOI phase. We will generate derived computational outputs:
\begin{itemize}
    \item Embedding-based protein similarity graph.
    \item Quantitative module recovery scores.
    \item PPI proximity scores to known disease genes.
    \item Ranked candidate proteins supported by both embedding and network evidence.
\end{itemize}

\section{Project Overview}

We use a simple pipeline that is easy to audit and explain.

\textbf{Planned workflow}
\begin{enumerate}
    \item \textbf{Map IDs across resources.} DrugCLIP, PPI, and disease resources use different identifier systems (for example UniProt IDs versus gene symbols). We first create a single crosswalk table so the same biological entity is represented consistently across all inputs. This step is critical because all downstream network joins depend on it; even a small mismatch rate can create false negatives or artificial sparsity.
    \item \textbf{Build an embedding similarity graph.} We represent each protein as a node and connect it to its top-$k$ nearest neighbors by cosine similarity in embedding space. The result is a kNN graph that converts high-dimensional vectors into a structure we can inspect and test directly. This graph is our operational definition of ``embedding-derived similarity'' and enables AD-vs-control comparisons.
    \item \textbf{Create AD and control seed neighborhoods.} We treat AD genes as primary seeds and matched non-AD genes as control seeds, then collect local embedding neighborhoods (1-hop; optionally 2-hop for sensitivity checks). This provides a fair side-by-side comparison under identical settings.
    \item \textbf{Compute region-specific support.} For each seed set, we score enrichment and specificity in entorhinal cortex, hippocampus, and temporal cortex (and relevant cell types). We summarize this as an IntegratedScore that combines expression strength, regional specificity, and network support.
    \item \textbf{Test module recovery and PPI proximity.} We measure overlap with AD-relevant modules and graph proximity to known AD genes, then compare AD seeds against control seeds and null models. This checks whether prioritized genes are not only close in embedding space, but also close to AD mechanisms in an independent network.
\end{enumerate}

\textbf{IntegratedScore definition (used for ranking).}
For each gene $g$, we compute
\[
    \mathrm{IntegratedScore}(g)=
    w_E \cdot Z_E(g) +
    w_S \cdot Z_S(g) +
    w_N \cdot Z_N(g),
\]
where $w_E+w_S+w_N=1$ and default weights are $w_E=0.40$, $w_S=0.30$, $w_N=0.30$.

\textbf{Component terms}
\begin{itemize}
    \item $Z_E(g)$: standardized expression strength in AD-vulnerable regions (entorhinal, hippocampus, temporal cortex), averaged across available donors/cell strata.
    \item $Z_S(g)$: standardized AD-region specificity, computed as expression contrast between AD-vulnerable regions and other cortical regions.
    \item $Z_N(g)$: standardized network support, combining (i) embedding-neighborhood AD enrichment and (ii) PPI proximity to known AD genes.
\end{itemize}
IntegratedScore is a transparent linear fusion of normalized signals, easy to ablate and sensitivity-test by changing weights. A higher score means a gene is simultaneously more expressed where AD pathology is prominent and better connected to known AD biology, making it a stronger mechanistic prioritization candidate.

\newpage
\textbf{Minimum controls and stress tests}

\textbf{Control 1: Label-shuffle test.}
We randomly permute disease labels while keeping graph structure fixed, then rerun the same module-recovery and proximity analyses. The purpose is to test whether the observed signal depends on true disease assignments. If performance remains high after shuffling, the original result is likely an artifact; if performance collapses, it supports disease-specific signal in the embeddings.

\textbf{Control 2: Random-embedding baseline.}
We replace DrugCLIP vectors with random vectors of the same dimension and scale, rebuild the kNN graph, and rerun all metrics. This establishes a floor for what purely geometric chance can produce. DrugCLIP should outperform this baseline clearly; otherwise, the claimed biological structure is not stronger than random embedding geometry.

\textbf{Control 3: Degree-matched random neighborhood baseline.}
For each seed, we sample random protein sets matched to neighborhood size (and matched degree distribution in the PPI graph), then compute the same overlap and proximity metrics. This prevents inflated performance due to hub bias and graph topology effects. It is a fairer baseline than uniform random sampling and is the main null model for significance testing.

\textbf{Control 4: Matched non-AD gene-set control.}
We run a fixed non-AD control gene set through the same full pipeline and compare distributions of IntegratedScore, module overlap, and PPI proximity to the AD set. This is the primary specificity control: if AD biology is truly captured, AD genes should rank higher and enrich more strongly in AD-vulnerable regions than matched controls.

\textbf{Stress test 1: kNN sensitivity sweep.}
We vary $k$ (for example 10, 25, 50) when constructing the embedding graph and check metric stability. Small $k$ can be too sparse; large $k$ can wash out local structure. A robust result should keep the same qualitative conclusion across a reasonable $k$ range.

\textbf{Stress test 2: PPI confidence-threshold sweep.}
We rerun analyses across multiple PPI confidence cutoffs (for example medium and high confidence edges only). This tests whether findings depend on noisy edges in the interaction graph. If conclusions hold under stricter PPI filtering, confidence in biological relevance increases.

\textbf{Primary deliverables}
\begin{itemize}
    \item Metric table comparing AD gene set vs non-AD controls across all baselines.
    \item Clear graph visualizations (embedding neighborhoods + PPI overlap).
    \item Ranked list of AD-prioritized proteins with IntegratedScore evidence columns.
    \item Reproducible code and compact error analysis.
\end{itemize}

\section{Impact and Applications}
If the answer is yes, we get evidence that DrugCLIP embeddings carry disease-mechanistic structure and can be used for network-informed target discovery. If the answer is no, we still learn something useful: current embeddings are not enough for this use case, and explicit network or spatial conditioning should be added next.

\bibliographystyle{unsrturl}
\bibliography{loi}

\end{document}
